Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-08-16 prop_letter.tex
prop_letter.tex After meeting...

A \emph{probability space} is a mathematical abstraction specifying
the necessary conditions for reasoning coherently about collections of
uncertain
events~\cite{Kolmogorov1950,544199,Griffiths2003,Grabisch2016}. In the
quantum case, the events of interest are specified by \emph{projection
  operators} $P$. These include the empty projector~$\mathbb{0}$,
the identity projector~$\mathbb{1}$, projectors of the form
$\proj{\phi}$ where $\ket{\phi}$ is a pure quantum state (an element
of a Hilbert space), sums of \emph{orthogonal} projectors $P_0$
and~$P_1$ with $P_0P_1=\mathbb{0}$, and products of \emph{commuting}
projectors $P_0$ and~$P_1$ with $P_0P_1=P_1P_0$. In a well-defined
quantum probability
space~\cite{10.2307/2308516,gleason1957,Redhead1987-REDINA,Maassen2010},
each event~$P_{i}$ is mapped to a probability~$\mu(P_{i})$ using a
probability measure~$\mu:\events\rightarrow[0,1]$, where $\events$
is the set of all projectors on a given Hilbert space, subject to the
following constraints: $\mu(\mathbb{0})=0$, $\mu(\mathbb{1})=1$,
$\mu\left(\mathbb{1}-P\right)=1-\mu\left(P\right)$, and for each pair
of \emph{orthogonal} projectors $P_{0}$ and $P_{1}$:
\begin{equation}
{\mu}\left(P_{0}+P_{1}\right)={\mu}\left(P_{0}\right)+{\mu}\left(P_{1}\right)\,.\label{eq:QuantumProbability-Addition}
\end{equation}

\noindent A conventional quantum probability measure can be easily
constructed if one knows the current state of the quantum system by
using the Born rule.  Specifically, for each pure normalized quantum
state $\ket{\phi}$, the Born rule induces a probability measure
$\mu_{\phi}^{\mathrm{B}}$ defined as
$\mu_{\phi}^{\mathrm{B}}(P)=\ip{\phi}{P\phi}$. The situation
generalizes to mixed states
$\rho = \sum_{j=1}^{N}q_{j}\proj{\phi_{j}}$, where
$\sum_{j=1}^{N}q_{j}=1$ in which case the generalized Born rule
induces a probability measure $\mu_{\rho}^{\mathrm{B}}$ defined as
$\mu_{\rho}^{\mathrm{B}}\left(P\right) = \Tr\left(\rho P\right) =
\sum_{j=1}^{N}
q_{j}\mu_{\phi_{j}}^{\mathrm{B}}\left(P\right)$~\cite{peres1995quantum,544199,Jaeger2007}.

As an example, consider a three-dimensional Hilbert space with
orthonormal basis $\{\ket{0},\ket{1},\ket{2}\}$ and an observable
$\mathbf{O}$ with spectral decomposition
$\mathbf{O} = \proj{0} + 2\,\proj{1} + 3\,\proj{2}$. Two fragments of
valid probability measures $\mu_1$ and $\mu_2$ that can be associated
with this space are: $\mu_1(\proj{0}) = 0.5$, $\mu_1(\proj{1}) = 0.5$,
$\mu_1(\proj{2}) = 0$, and $\mu_2(\proj{0}) = 0.5$,
$\mu_2(\proj{1}) = 0$, $\mu_2(\proj{2}) = 0.5$. By the Born rule, the
first probability measure corresponds to the quantum system being in
the pure state $(\ket{0}+\alpha\ket{1})/\sqrt{2}$ and the second
corresponds to the quantum system being in the pure state
$(\ket{0}+\alpha\ket{2})/\sqrt{2}$, where in both cases
$|\alpha|=1$. The expectation value of the observable $\mathbf{O}$ in
the first case is 1.5 and in the second it is 2.

@@

Our generalization of quantum probability measures to QIVPMs will
allow us to strengthen one of the fundamental theorems in of quantum
information: the Kochen-Specker
theorem~\cite{PeresRon1988,peres1995quantum}. Our stronger version of
the theorem will suggest a resolution to the debate initiated by Meyer
and Mermin on the relevance of the Kochen-Specker to experimental, and
hence finite-precision, quantum
measurements~\citep{PhysRevLett.83.3751,Mermin1999,Kent1999,HavlicekKrennSummhammerSvozil2001,SimonBruknerZeilinger2001,Cabello2002,Larsson2002,Appleby2002,Peres2003,BarrettKent2004,Appleby_2005,Spekkens2005,GuehneKleinmannCabelloEtAl2010,MazurekPuseyKunjwalEtAl2016}.
. Specifically, the original Kochen-Specker theorem is formulated
under the assumption that all observables of a quantum mechanical
system have \emph{definite values at all
  times}~\cite{Held2016}. Our interval-valued framework will allow
us to state, and prove, a stronger version of the theorem that holds
even if the observables have values that are only definite up to some
specific parameter that quantifies experimental uncertainty. Put
differently, our stronger statement and theorem quantifies Mermin's
qualitative argument:
\begin{quote}
  \ldots although the outcomes deduced from such imperfect
  measurements will occasionally differ dramatically from those
  allowed in the ideal case, if the misalignment is very slight, the
  statistical distribution of outcomes will differ only slightly from
  the ideal case~\citep{Mermin1999}.
\end{quote}

@@

\begin{definition} Given a Hilbert space $\Hilb$ of dimension~$d\ge3$,
  for every QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}$, any
  observable~$\mathbf{O}$ with spectral decomposition
  $\mathbf{O}=\sum_{i=1}^{N}\lambda_{i}P_{i}$, and the minimal
  subspace of events $\events'$ containing
  $\left\{ P_{i}\right\} _{i=1}^{N}$, we have
\begin{equation}
\expval{\mathbf{O}}_{\bar{\mu}}=\left[\min_{\rho\in\coreBorn\left(\bar{\mu},\events'\right)}\expval{\mathbf{O}}_{\rho},\max_{\rho\in\coreBorn\left(\bar{\mu},\events'\right)}\expval{\mathbf{O}}_{\rho}\right]\,.\label{eq:quantum-interval-expectation-core}
\end{equation}
\end{definition}

@@

\begin{thm}[0-deterministic variant of Kochen-Specker] 
\label{thm:Kochen-Specker}
Given a Hilbert space $\Hilb$ of dimension~$d\ge3$, there is no
0-deterministic
measure~$\mu^{\mathrm{D}}:\events\rightarrow\left\{ 0,1\right\}$.
\end{thm}

>>>>

A \emph{probability space} is a mathematical abstraction specifying
the necessary conditions for reasoning coherently about collections of
uncertain
events~\cite{Kolmogorov1950,544199,Griffiths2003,Grabisch2016}. In the
quantum case, the events of interest are specified by \emph{projection
  operators} $P$. These include the empty projector~$\mathbb{0}$,
the identity projector~$\mathbb{1}$, projectors of the form
$\proj{\phi}$ where $\ket{\phi}$ is a pure quantum state (an element
of a Hilbert space), sums of \emph{orthogonal} projectors $P_0$
and~$P_1$ with $P_0P_1=\mathbb{0}$, and products of \emph{commuting}
projectors $P_0$ and~$P_1$ with $P_0P_1=P_1P_0$. In a well-defined
quantum probability
space~\cite{10.2307/2308516,gleason1957,Redhead1987-REDINA,Maassen2010},
each event~$P_{i}$ is mapped to a probability~$\mu(P_{i})$ using a
probability measure~$\mu:\events\rightarrow[0,1]$, where $\events$
is the set of all projectors on a given Hilbert space, subject to the
following constraints: $\mu(\mathbb{0})=0$, $\mu(\mathbb{1})=1$,
$\mu\left(\mathbb{1}-P\right)=1-\mu\left(P\right)$, and for each pair
of \emph{orthogonal} projectors $P_{0}$ and $P_{1}$:
\begin{equation}
{\mu}\left(P_{0}+P_{1}\right)={\mu}\left(P_{0}\right)+{\mu}\left(P_{1}\right)\,.\label{eq:QuantumProbability-Addition}
\end{equation}
Given an observable~$\mathbf{O}$ with spectral decomposition
$\mathbf{O}=\sum_{i=1}^{N}\lambda_{i}P_{i}$ one can define the 
expectation value of that observable as
\begin{eqnarray}
\langle \mathbf{O} \rangle_{\mu}=
\sum_{i=1}^{N}\lambda_{i}\mu(P_{i}) .
\end{eqnarray}

A conventional quantum probability measure can be easily
constructed if one knows the current state of the quantum system by
using the Born rule.  Specifically, for each pure normalized quantum
state $\ket{\phi}$, the Born rule induces a probability measure
$\mu_{\phi}^{\mathrm{B}}$ defined as
$\mu_{\phi}^{\mathrm{B}}(P)=\ip{\phi}{P\phi}$. The situation
generalizes to mixed states
$\rho = \sum_{j=1}^{N}q_{j}\proj{\phi_{j}}$, where
$\sum_{j=1}^{N}q_{j}=1$ in which case the generalized Born rule
induces a probability measure $\mu_{\rho}^{\mathrm{B}}$ defined as
$\mu_{\rho}^{\mathrm{B}}\left(P\right) = \Tr\left(\rho P\right) =
\sum_{j=1}^{N}
q_{j}\mu_{\phi_{j}}^{\mathrm{B}}\left(P\right)$~\cite{peres1995quantum,544199,Jaeger2007}.

As an example, consider a three-dimensional Hilbert space with
orthonormal basis $\{\ket{0},\ket{1},\ket{2}\}$ and an observable
$\mathbf{O}$ with spectral decomposition
$\mathbf{O} = \proj{0} + 2\,\proj{1} + 3\,\proj{2}$. Two fragments of
valid probability measures $\mu_1$ and $\mu_2$ that can be associated
with this space are: $\mu_1(\proj{0}) = 0.5$, $\mu_1(\proj{1}) = 0.5$,
$\mu_1(\proj{2}) = 0$, and $\mu_2(\proj{0}) = 0.5$,
$\mu_2(\proj{1}) = 0$, $\mu_2(\proj{2}) = 0.5$. By the Born rule, the
first probability measure corresponds to the quantum system being in
the pure state $\ket{\phi_1}=(\ket{0}+\alpha\ket{1})/\sqrt{2}$ and the second
corresponds to the quantum system being in the pure state
$\ket{\phi_2}=(\ket{0}+\alpha\ket{2})/\sqrt{2}$, where in both cases
$|\alpha|=1$. The expectation value of the observable $\mathbf{O}$, 
$\langle \mathbf{O} \rangle_{\mu_{1,2}}$,in
the first case is 1.5 and in the second it is 2.

@@

Our generalization of quantum probability measures to QIVPMs will
allow us to strengthen one of the fundamental theorems in of quantum
information: the Kochen-Specker
theorem~\cite{PeresRon1988,peres1995quantum}. Our stronger version of
the theorem will suggest a resolution to the debate initiated by Meyer
and Mermin on the relevance of the Kochen-Specker to experimental, and
hence finite-precision, quantum
measurements~\citep{PhysRevLett.83.3751,Mermin1999,Kent1999,HavlicekKrennSummhammerSvozil2001,SimonBruknerZeilinger2001,Cabello2002,Larsson2002,Appleby2002,Peres2003,BarrettKent2004,Appleby_2005,Spekkens2005,GuehneKleinmannCabelloEtAl2010,MazurekPuseyKunjwalEtAl2016}.
Specifically, the original Kochen-Specker theorem is formulated
under the assumption that all observables of a quantum mechanical
system have \emph{definite values at all
  times}~\cite{Held2016}. Our interval-valued framework will allow
us to state, and prove, a stronger version of the theorem that holds
even if the observables have values that are only definite up to some
specific parameter that quantifies experimental uncertainty. Put
differently, our stronger statement and theorem quantifies Mermin's
qualitative argument:
\begin{quote}
  \ldots although the outcomes deduced from such imperfect
  measurements will occasionally differ dramatically from those
  allowed in the ideal case, if the misalignment is very slight, the
  statistical distribution of outcomes will differ only slightly from
  the ideal case~\citep{Mermin1999}.
\end{quote}

@@

\begin{thm} Given a Hilbert space $\Hilb$ of dimension~$d\ge3$,
  for every QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}$, and the minimal
  subspace of events $\events'$ containing
  $\left\{ P_{i}\right\} _{i=1}^{N}$, then
\begin{equation}
\expval{\mathbf{O}}_{\bar{\mu}}=\left[\min_{\rho\in\coreBorn\left(\bar{\mu},\events'\right)}\expval{\mathbf{O}}_{\rho},\max_{\rho\in\coreBorn\left(\bar{\mu},\events'\right)}\expval{\mathbf{O}}_{\rho}\right]\,.\label{eq:quantum-interval-expectation-core}
\end{equation}
\end{thm}

@@

\begin{thm}
\label{thm:Kochen-Specker} {\rm [0-deterministic variant of Kochen-Specker]}
Given a Hilbert space $\Hilb$ of dimension~$d\ge3$, there is no
0-deterministic
measure~$\mu^{\mathrm{D}}:\events\rightarrow\left\{ 0,1\right\}$.
\end{thm}
